Assessing Replication via P-values

### 2018-06-25 draft_mcnemars_ttest.Rmd
appendix added with some light explainations of the tables.

The null hypothesis for McNemar's test states that the marginal probabilities for each outcome are the same. That is, $H_0: p_{11} + p_{10} = p_{11} + p_{01}$ where the $P$'s are the probabilities of occurrence in cells with the corresponding labels. The test statistic is given by $\chi^2 = \frac{(n_{10} - n_{01})^2}{n_{10} + n_{01}}$. Under the null hypothesis, the test statistic follow a chi-squared distribution with one degree of freedom. The null hypothesis is reject for large values of the test statistic. The results of the OSF article are summarized in the contingency table below. 97% of original findings were significant compared to 36% of replication findings. The authors found a test statistic of $X^2 = 59.06$, rejected the null hypothesis, and concluded that the findings from the original studies did not replicate.

@@

Assuming the true null hypothesis for replication is $H_0: \theta_{1j} = \theta_{2j}$, a Type I error occurs when the original and replication studies differ in their power to detect their common effect size. In this case, McNemar's test will almost always reject its null hypothesis leading to conclusions of irreplicability despite having identical true effects. The power of each of the $2k$ original and replication studies is determined by the true effect size $(\theta_{ij})$, the sample size $(n_{ij})$, and the population variance $(\sigma^2)$. If the original and replication studies have the same true effect size and population variance, we can let their powers differ by increasing the sample size of the replication studies. As $n_{2j}$ grows large, $\frac{|T_{2j}|}{\sqrt{v_{2j}}}$ will also grow large. If we assume that original and replication studies are independent then,

@@

Like McNemar's test, the power of the t-test to detect if the \textit{p}-values of the original studies are different from those of the replication studies depends upon the power of both the original and replication studies. Consider a single study (either original or replication), with the null hypothesis $H_0:\theta_i = 0$. The finding from this study is significant if $\frac{|T_{i}|}{\sqrt{v_i}} \geq 1.96$ which is equivalent to $p_i \leq 0.05$ since $p_i = \Phi^{-1} \Big ( \frac{|T_i|}{\sqrt{v_i}} \Big)$ where $\Phi^{-1}$ is the standard normal quantile function.. Assuming that the variance is known for such a study and is $\frac{4}{n}$ where $n$ is the sample size then, the power to detect if $\theta_i$ is statistically different from 0 depends on the estimate $T_i$ and the sample size $n$. Extending this to the test statistic for the t-test, we note that the mean of the paired difference of \textit{p}-values can be written as

@@

 Given that this mean takes into account both the estimate and sample size of the original and replication studies, the power to detect a difference in \textit{p}-values depends on the power of both the original and replication studies.

Again, we consider the Type I and Type II error for the t-test. A Type I error occurs when the effect sizes from the original and replication studies are equal but the t-test concludes that their \textit{p}-values are significant different from one another. This is likely to occur when the power of the original and replication studies differ. In a simple thought experiment, we consider a single study pair. The t-test used to compare the original and replication findings in this pair would have a test statistic of $\frac{{p_1-p_2}}{\sqrt{V_{p_1-p_2}}}$. If we fix the parameters of the original study and let the sample size of the replication studies tend to $\infty$, the variance of the replication estimate will tend to 0 and, as a result, $p_2$ will also tend to 0. This would not be unusual in a real-world appication as researchers often increase replication sample sizes to get a more precise estimate of the original finding. In this scenario, the t-test stastic depends only on the \textit{p}-value of the original study $\Big( \frac{p_1}{\sqrt{V_{p_1}}}\Big)$ and the error rate tends to 1. In other words, the t-test will rarely conclude that the two \textit{p}-values are similar to one another. Another way to think about this error rate is to consider the case where the original and replication test statistics differ (i.e. $\frac{T_1}{\sqrt{V_1}} \neq \frac{T_2}{\sqrt{V_2}}$). If the original and replication studies have different power to detect the same effect, the numerators of these statistics will be similar, but the sample sizes and variances will differ. As a result, the \textit{p}-values from the original and replication studies will be quite different, and once again the Type I error rate will tend towards 0.80.

@@

In data simulations to test our analysis of the Type I error rate, we fix the effect size of the original and replication studies at either a small, medium, or large effect. As is convention, we consider $\theta_{ij} = 0.2$ to be a small effect, $\theta_{ij} = 0.5$ to be a medium effect, and $\theta_{ij} = 0.8$ to be a large effect. For each of these effect size specifications, we then consider the case where (1) the original study is powered at 40% and the replication study is powered at 60%, (2) the original study is powered at 60% and the replication study is powered at 80%, and (3) the original study is powered at 40% and the replication study is powered at 80%. These power specifications are consistent with the range observed in the OSF data. For each of these data specifications, we run 10,000 simulations and calculate the proportion in which the t-test or McNemar's test reject the null hypothesis in favor of the alternative. The results of these simulations are displayed in Table 1. As expected, the probability of rejecting a true null hypothesis ranges from 0.8 to 1.0 for the two tests.

\begin{table}[h!]
\centering
\begin{tabular}{|l | c | c | c |}
\hline\hline
& \textbf{Power = 40/60} & \textbf{Power = 60/80} & \textbf{Power = 40/80} \\
\hline
$\theta =$ \textbf{0.2} & & & \\
\hspace{3mm} McNemar's & 0.808 (0.01) & 0.873 (0.01) & 0.999 (0.0003) \\
\hspace{3mm} T test & 0.849 (0.01) & 0.825 (0.01) & 0.999 (0.0002)\\
\hline
$\theta =$ \textbf{0.5} & & & \\
\hspace{3mm} McNemar's & 0.808 (0.01) & 0.874 (0.01) & 0.999 (0.0002)  \\
\hspace{3mm} T test & 0.851 (0.01) & 0.826 (0.01) & 0.999 (0.0002) \\
\hline
$\theta =$ \textbf{0.8} & & & \\
\hspace{3mm} McNemar's & 0.804 (0.01) & 0.874 (0.01) & 0.999 (0.0001) \\
\hspace{3mm} T test & 0.851 (0.01) & 0.824 (0.01) & 1.000 (0.0000)\\
\hline\hline
\end{tabular}
\caption{Probability of rejecting a true null hypothesis for the \textit{t} test and McNemar's test. Standard errors are reported in  parentheses and rounded to the first significant digit.}
\end{table}

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l | c | c | c |}
\hline\hline
& $\theta =$ \textbf{0.2/0.5} & $\theta=$ \textbf{0.5/0.8} & $\theta=$ \textbf{0.2/0.8} \\
\hline
\textbf{40\%\ Power} & & & \\
\hspace{3mm} McNemar's & 0.049 (0.007) & 0.050 (0.007) & 0.049 (0.007) \\
\hspace{3mm} T test & 0.050 (0.007) & 0.051 (0.007) & 0.050 (0.007)\\
\hline
\textbf{60\%\ Power} & & & \\
\hspace{3mm} McNemar's & 0.050 (0.007) & 0.050 (0.007) & 0.049 (0.007) \\
\hspace{3mm} T test & 0.050 (0.006) & 0.049(0.008) & 0.049 (0.006)\\
\hline
\textbf{80\%\ Power} & & & \\
\hspace{3mm} McNemar's & 0.051 (0.008) & 0.051 (0.007) & 0.051 (0.008) \\
\hspace{3mm} T test &  0.048 (0.008) & 0.048 (0.007) & 0.048 (0.007)\\
\hline\hline
\end{tabular}
\caption{Probability of rejecting a false null hypothesis for the \textit{t} test and McNemar's test. Standard errors are reported in paraentheses and rounded to the first significant digit.}
\end{table}

>>>>

The null hypothesis for McNemar's test states that the marginal probabilities for each outcome are the same. That is, $H_0: p_{11} + p_{10} = p_{11} + p_{01}$ where the $p$'s are the probabilities of occurrence in cells with the corresponding labels. The test statistic is given by $\chi^2 = \frac{(n_{10} - n_{01})^2}{n_{10} + n_{01}}$. Under the null hypothesis, the test statistic follow a chi-squared distribution with one degree of freedom. The null hypothesis is reject for large values of the test statistic. The results of the OSF article are summarized in the contingency table below. 97% of original findings were significant compared to 36% of replication findings. The authors found a test statistic of $X^2 = 59.06$, rejected the null hypothesis, and concluded that the findings from the original studies did not replicate.

@@

Assuming the true null hypothesis for replication is $H_0: \theta_{1j} = \theta_{2j}$, a Type I error occurs when the original and replication studies differ in their power to detect their common effect size. In this case, McNemar's test will almost always reject its null hypothesis leading to conclusions of irreplicability despite having identical true effects. The power of each of the $2k$ original and replication studies is determined by the true effect size $(\theta_{ij})$, the sample size $(n_{ij})$, and the population variance $(\sigma^2)$. If the original and replication studies ahve the same true effect size and population variance, we can let their powers differ by increasing the sample size of the replication studies. As $n_{2j}$ grows large, $\frac{|T_{2j}|}{\sqrt{v_{2j}}}$ will also grow large. If we assume that original and replication studies are independent then,

@@

Like McNemar's test, the power of the t-test to detect if the \textit{p}-values of the original studies are different from those of the replication studies depends upon the power of both the original and replication studies. Consider a single study (either original or replication), with the null hypothesis $H_0:\theta_i = 0$. The finding from this study is significant if $\frac{|T_{i}|}{\sqrt{v_i}} \geq 1.96$ which is equivalent to $p_i \leq 0.05$ since $p_i = \Phi^{-1} \Big ( \frac{|T_i|}{\sqrt{v_i}} \Big)$. Assuming that the variance is known for such a study and is $\frac{4}{n}$ where $n$ is the sample size then, the power to detect if $\theta_i$ is statistically different from 0 depends on the estimate $T_i$ and the sample size $n$. Extending this to the test statistic for the t-test, we note that the mean of the paired difference of \textit{p}-values can be written as

@@

where $\Phi^{-1}$ is the standard normal quantile function. Given that this mean takes into account both the estimate and sample size of the original and replication studies, the power to detect a different in \textit{p}-values depends on the power of both the original and replication studies.

Again, we consider the Type I and Type II error for the t-test. A Type I error occurs when the effect sizes from the original and replication studies are equal but the t-test concludes that their \textit{p}-values are significant different from one another. This is likely to occur when the power of the original and replication studies differ. In a simple thought experiment, we consider a single study pair. The t-test to compare the original and replication findings in this pair would have a test statistic of $\frac{{p_1-p_2}}{\sqrt{V_{p_1-p_2}}}$. If we fix the parameters of the original study and let the sample size of the replication studies tend to $\infty$, the variance of the replication estimate will tend to 0 and, as a result, $p_2$ will also tend to 0. This would not be unusual in a real-world appication as researchers often increase replication sample sizes to get a more precise estimate of the original finding. In this scenario, the t-test stastic depends only on the \textit{p}-value of the original study $\Big( \frac{p_1}{\sqrt{V_{p_1}}}\Big)$ and the error rate tends to 1. In other words, the t-test will rarely conclude that the two \textit{p}-values are similar to one another. Another way to think about this error rate is to consider the case where the original and replication test statistics differ (i.e. $\frac{T_1}{\sqrt{V_1}} \neq \frac{T_2}{\sqrt{V_2}}$). If the original and replication studies have different power to detect the same effect, the numerators of these statistics will be similar, but the sample sizes and variances will differ. As a result, the \textit{p}-values from the original and replication studies will be quite different, and once again the Type I error rate will tend towards 0.80.

@@

In data simulations to test our analysis of the Type I error rate, we fix the effect size of the original and replication studies at either a small, medium, or large effect. As is convention, we consider $\theta_{ij} = 0.2$ to be a small effect, $\theta_{ij} = 0.5$ to be a medium effect, and $\theta_{ij} = 0.8$ to be a large effect. For each of these effect size specifications, we then consider the case where (1) the original study is powered at 40% and the replication study is powered at 60%, (2) the original study is powered at 60% and the replication study is powered at 80%, and (3) the original study is powered at 40% and the replication study is powered at 80%. These power specifications are consistent with the range observed in the OSF data. For each of these data specifications, we run 10,000 simulations and calculate the proportion in which the t-test or McNemar's test reject the null hypothesis in favor of the alternatie. The results of these simulations are displayed in Table 1. As expected, the probability of rejecting a true null hypothesis ranges from 0.8 to 1.0 for the two tests.

\begin{table}[h!]
\centering
\begin{tabular}{|l | c | c | c |}
\hline\hline
& \textbf{Power = 40/60} & \textbf{Power = 60/80} & \textbf{Power = 40/80} \\
\hline
$\theta =$ \textbf{0.2} & & & \\
\hspace{3mm} McNemar's & 0.808 (0.01) & 0.873 (0.01) & 0.999 (0.0003) \\
\hspace{3mm} T test & 0.849 (0.01) & 0.825 (0.01) & 0.999 (0.0002)\\
\hline
$\theta =$ \textbf{0.5} & & & \\
\hspace{3mm} McNemar's & 0.808 (0.01) & 0.874 (0.01) & 0.999 (0.0002)  \\
\hspace{3mm} T test & 0.851 (0.01) & 0.826 (0.01) & 0.999 (0.0002) \\
\hline
$\theta =$ \textbf{0.8} & & & \\
\hspace{3mm} McNemar's & 0.804 (0.01) & 0.874 (0.01) & 0.999 (0.0001) \\
\hspace{3mm} T test & 0.851 (0.01) & 0.824 (0.01) & 1.000 (0.0000)\\
\hline\hline
\end{tabular}
\caption{Probability of rejecting a true null hypothesis for the \textit{t} test and McNemar's test. Standard errors are reported in  parentheses and rounded to one significant digit.}
\end{table}

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l | c | c | c |}
\hline\hline
& $\theta =$ \textbf{0.2/0.5} & $\theta=$ \textbf{0.5/0.8} & $\theta=$ \textbf{0.2/0.8} \\
\hline
\textbf{40\%\ Power} & & & \\
\hspace{3mm} McNemar's & 0.049 (0.007) & 0.050 (0.007) & 0.049 (0.007) \\
\hspace{3mm} T test & 0.050 (0.007) & 0.051 (0.007) & 0.050 (0.007)\\
\hline
\textbf{60\%\ Power} & & & \\
\hspace{3mm} McNemar's & 0.050 (0.007) & 0.050 (0.007) & 0.049 (0.007) \\
\hspace{3mm} T test & 0.050 (0.006) & 0.049(0.008) & 0.049 (0.006)\\
\hline
\textbf{80\%\ Power} & & & \\
\hspace{3mm} McNemar's & 0.051 (0.008) & 0.051 (0.007) & 0.051 (0.008) \\
\hspace{3mm} T test &  0.048 (0.008) & 0.048(.007) & 0.048 (0.007)\\
\hline\hline
\end{tabular}
\caption{Probability of rejecting a false null hypothesis for the \textit{t} test and McNemar's test. Standard errors are reported in paraentheses and rounded to the first significant digit.}
\end{table}
